Carryover effects in free recall reveal how prior experiences influence memories of new experiences

### 2022-12-27 paper/main.tex
backing up progress on intro
backing up progress to intro

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers, super]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}

@@

% the role of context and prior experience in memory
Experience is subjective: different people who encounter identical physical experiences
can take away very different meanings and memories.  One reason is that our subjective
experiences in the moment are shaped in part the idiosyncratic prior experiences, memories,
goals, thoughts, expectations, and emotions that we bring with us into the present moment.
These factors collectively define a \textit{context} for our experiences~\citep{Mann20}.

% situation models: forming expectations, predicting ambiguous future experiences
The contexts we encounter help us to construct \textit{situation models}~\citep{RangRitc12, MannEtal15}
or \textit{schemas}~\citep{MasiEtal22, BaldEtal18} that describe how experiences are likely to unfold
based on our prior experiences with similar contextual cues.  For example, when we enter a sit-down restaurant,
we might expect to be seated at a table, given a menu, and served food.  Priming someone
to expect a particular situation or context can also influence how they resolve potentail ambiguities
in their ongoing experiences, including ambiguous movies and narratives~\citep{YeshEtal17}.

%NOTE: should we start with the naturalistic framing?  or would it be more effective and direct to
% jump right into free recall?  i've added some potentially useful text to boneyard.tex that could be
% pasted in with some tweaking.  some of the above text could also be moved to the discussion section.

% gap between "classic" free recall tasks and naturalistic (or real-world) memory tasks
Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in
part by substantial differences in how we study these processes.  Situation models
and schemas are most often studied using ``naturalistic'' 
stimuli such as narratives and movies~\citep{CITES}.  In contrast, our
understanding of how we organize our memories has been most widely studied
using more traditional paradigms like free recall of random word lists~\citep{Kaha12}.
Because random word lists are unstructured by design, it is not clear if or how non-trivial
situation models might apply to these stimuli.

@@

>>>>

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[font=small,labelfont=bf]{caption}
\usepackage{geometry}
\usepackage[sort&compress, numbers, super]{natbib}
\usepackage{pxfonts}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage{hyperref}
\usepackage{lineno}

@@

% the role of context and prior experience in memory

Experience is subjective: different people who encounter identical physical
experiences can take away very different meanings and memories. One reason is
that our subjective experiences in the moment are shaped in part the
idiosyncratic prior experiences, memories, goals, thoughts, expectations, and
emotions that we bring with us into the present moment. These factors
collectively define a \textit{context} for our experiences~\citep{Mann20}. %
situation models: forming expectations, predicting ambiguous future experiences
The contexts we encounter help us to construct \textit{situation
models}~\citep{RangRitc12, MannEtal15} or \textit{schemas}~\citep{MasiEtal22,
BaldEtal18} that describe how experiences are likely to unfold based on our
prior experiences with similar contextual cues. For example, when we enter a
sit-down restaurant, we might expect to be seated at a table, given a menu, and
served food. Priming someone to expect a particular situation or context can
also influence how they resolve potentail ambiguities in their ongoing
experiences, including ambiguous movies and narratives~\citep{YeshEtal17}.

% gap between "classic" free recall tasks and naturalistic (or real-world)
% memory tasks

Our understanding of how we form situation models and schemas, and how they
interact with our subjective experiences and memories, is constrained in part
by substantial differences in how we study these processes. Situation models
and schemas are most often studied using ``naturalistic'' stimuli such as
narratives and movies~\citep{ZwaaEtal95,ZwaaRadv98, NastEtal20}. In contrast,
our understanding of how we organize our memories has been most widely studied
using more traditional paradigms like free recall of random word
lists~\citep{Kaha12}. In free recall, participants study lists of items and are
instructed to recall the items in any order they choose. The orders in which
words come to mind can provide insights into how participants have organized
their memories of the studied words. Because random word lists are unstructured
by design, it is not clear if or how non-trivial situation models might apply
to these stimuli. Nevertheless, there are \textit{some} commonalities between
memory for word lists and memory for real-world experiences.

Like remembering real-world experiences, remembering words on a studied list
requires distinguishing the current list from the rest of one's experience. To
model this fundamental memory capability, cognitive scientists have posited the
existence of a special representation, called \emph{context}, that is
associated with each list. According to early
theories~\citep[e.g.][]{Este55a,AndeBowe72} context representations are
composed of many features which fluctuate from moment to moment, slowly
drifting through a multidimensional feature space. During recall, this
representation forms part of the retrieval cue, enabling us to distinguish list
items from non-list items. Understanding the role of context in memory
processes is particularly important in self-cued memory tasks, such as
\textit{free recall}, where the retrieval cue is ``context'' itself.

Over the past half-century, context-based models have enjoyed impressive
success at explaining many stereotyped behaviors observed during free recall
and other list-learning tasks~\citep{Este55a, RaaiShif80, GlenEtal83,
HowaKaha02a, SiroEtal05, KimbEtal07, PolyKaha08, PolyEtalTulv, SedeEtal08,
PolyEtal09, ShanHowa12}. These phenomena include the well-known recency and
primacy effects (superior recall of items from the end and, to a lesser extent,
from the beginning of the study list), as well as semantic and temporal
clustering effects~\citep{KahaEtal08}. The contiguity effect is an example of
temporal clustering, which is perhaps the dominant form of organization in free
recall. This effect can be seen in the tendency for people to successively
recall items that occupied neighboring positions in the study list. For
example, if a list contained the sub-sequence \textsc{``absence hollow pupil''}
and the participant recalls the word \textsc{``hollow''}, it is far more likely
that the next response will be either \textsc{``pupil''} or
\textsc{``absence''} than some other list item~\citep{Kaha96}. In addition,
there is a strong forward bias in the contiguity effect: subjects make forward
transitions (i.e., \textsc{``hollow''} followed by \textsc{``pupil''}) about
twice as often as they make backward transitions, despite an overall tendency
to begin recall at the end of the list. There are also striking effects of
semantic clustering~\citep{RomnEtal93, Bous53, BousEtal54, JenkRuss52,
MannKaha12}, whereby the recall of a given item is more likely to be followed
by recall of a similar or related item than a dissimilar or unrelated one.

In general, people organize memories for words along a wide variety of stimulus
dimensions. As captured by models like the \textit{Context Maintenance and
Retrieval Model}~\citep{PolyEtal09}, the stimulus features associated with each
word (e.g.\ the word's meaning, font size, font color, location on the screen,
size of the object the word represents, etc.) are incorporated into the
participant's mental context representation~\citep{SmitVela01, MannEtal15, Mann20}.
During a memory test, any of these features may serve as a memory cue, which in
turn leads the participant to recall in succession words that share stimulus
features.

% link clustering to schemas...

These clustering behaviors in ....

@@

\begin{figure} 
    \begin{center}
        \includegraphics[width=0.4 \textwidth]{figures/FRFR} \end{center} \vspace{-16pt}

        \caption{\textbf{Feature-rich free recall.} After studying lists
comprised of words that vary along several feature dimensions, participants
verbally recall words in any order (microphone icon).}

        \label{fig:exp}
\end{figure}
